---
output:
  html_document: default
---


## 1 load libraries and breach report data

```{r}
library(readr)
library(tidyr)
library(dplyr)
 breach_report <- read_csv("~/Downloads/breach_report.csv")
# View(breach_report)this throws an error when i knit not sure why.  this is the error Quitting from lines 22-30 (Capstone_RMarkdown.Rmd) 
#Error in check_for_XQuartz() : 
 # X11 library is missing: install XQuartz from xquartz.macosforge.org
#Calls: <Anonymous> ... withVisible -> eval -> eval -> View -> check_for_XQuartz
#Execution halted
```
## 2 review and clean data
review states for unique id...would expect 50 states but identified 53.  upon review, realized DC is in population as a territory as is Puerto Rico (PR)  Reviewed NA findings and these appear to be Puerto Ric0.  
rename state column as US States and Territories to be more clear.  replaced is.na findings with PR
review each column for blank or NA values

```{r}

summary(breach_report)
unique(breach_report$State)
breach_report %>% filter(State=="DC")
breach_report %>% filter(is.na(State))
breach_report %>% filter(State=="PR")
breach_report<- rename(breach_report, State_Territory = State)
breach_report$State_Territory[is.na(breach_report$State_Territory)]<-"PR"
```
## continue to review and clean as needed.  review each column or blank or NA values
```{r}
breach_report %>% filter(`Name of Covered Entity`==""|is.na(`Name of Covered Entity`))
```
