---
output:
  html_document: default
---


## 1 load libraries and breach report data

```{r}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
 breach_report <- read.csv("~/Downloads/breach_report.csv")
 head(breach_report,3)


 # use summary or head to display data

```
## 2 review and clean data
review states for unique id...would expect 50 states but identified 53.  upon review, realized DC is in population as a territory as is Puerto Rico (PR)  Reviewed NA findings and these appear to be Puerto Ric0.  
rename state column as US States and Territories to be more clear.  replaced is.na findings with PR
review each column for blank or NA values

```{r}

summary(breach_report,3)
unique(breach_report$State)
breach_report %>% filter(State=="DC")
breach_report %>% filter(is.na(State))
breach_report %>% filter(State=="PR")
breach_report<- rename(breach_report, State.Territory = State)
breach_report$State.Territory[is.na(breach_report$State.Territory)]<-"PR"
```
## continue to review and clean as needed.  review each column or blank or NA values
# check to see if there are any blank values 
```{r}
breach_report %>% filter(`Type.of.Breach`=="") %>% filter(`Web.Description`=="") %>% filter(`Name.of.Covered.Entity`=="") %>% filter(`Individuals.Affected`=="") %>% filter(`Location.of.Breached.Information`=="") %>% filter(`Breach.Submission.Date`=="") %>% filter(`Business.Associate.Present`=="") %>% filter(`Covered.Entity.Type`=="") %>% filter(`Covered.Entity.Type`=="") 
```
#no blank values.  
#convert date column from character to date format m/d/y
```{r}
breach_report$`Breach.Submission.Date`<- as.Date(breach_report$`Breach.Submission.Date`,"%m/%d/%Y")
```
#review remaining columns.  replace NA values with "Unknown" for character values, 01/01/1970 for dates and 0 for numeric values(still deciding on the no value for numerics)
```{r}
breach_report %>% filter(Name.of.Covered.Entity ==""|is.na(Name.of.Covered.Entity))
# head or summary for breach_report
# summary(breach_report$Name of Covered Entity,3)
breach_report %>% filter(`Individuals.Affected`==""|is.na(`Individuals.Affected`))
# some are listed as NA so number is unknown.  replacing with values of 0
#reconsidering....populating with 0 may not be correct....
#breach_report$`Individuals.Affected`[is.na(breach_report$`Individuals.Affected`)]<-0
# re run check for blank or na to make sure was corrected
breach_report %>% filter(`Individuals.Affected`==""|is.na(`Individuals.Affected`))
# check for na or blank values for breached info  replacing with unknown
breach_report %>% filter(`Location.of.Breached.Information`==""|is.na(`Location.of.Breached.Information`))
breach_report$`Location.of.Breached.Information`[is.na(breach_report$`Location.of.Breached Information`)]<-"unknown"
breach_report %>% filter(`Location.of.Breached.Information`==""|is.na(`Location.of.Breached.Information`))
breach_report$`Type.of.Breach`[is.na(breach_report$`Type.of.Breach`)]<-"Unknown"
breach_report$`Name.of.Covered Entity`[is.na(breach_report$`Name.of.Covered.Entity`)]<-"Unknown"
breach_report$`Covered.Entity.Type`[is.na(breach_report$`Covered.Entity.Type`)]<-"Unknown"
#breach_report$`Breach Submission Date`[is.na(breach_report$`Breach Submission Date`)]<- "01/01/1070"
breach_report$`Type.of.Breach`[is.na(breach_report$`Type.of.Breach`)]<-"Unknown"
breach_report$`Business.Associate.Present`[is.na(breach_report$`Business Associate Present`)]<-"Unknown"
breach_report$`Web.Description`[is.na(breach_report$`Web Description`)]<-"Unknown"
head(breach_report,3)

```
##start plotting
```{r}
#following gives me an error can we talk through?
#vapply(strsplit(breach_report$Type.of.Breach,",",fixed = TRUE),"[","",1)
breach_report$Type.of.Breach<-sub(pattern = "^H.*",replacement = "Hacking/IT Incident" ,x=breach_report$Type.of.Breach)
breach_report$Type.of.Breach<-sub(pattern = "^H.*",replacement = "Hacking/IT Incident" ,x=breach_report$Type.of.Breach)
breach_report$Type.of.Breach<-sub(pattern = "^H.*",replacement = "Hacking/IT Incident" ,x=breach_report$Type.of.Breach)
breach_report$Type.of.Breach<-sub(pattern = "^U.*",replacement = "Unauthorized Access/Disclosure", x=breach_report$Type.of.Breach)
 breach_report$Type.of.Breach<-sub(pattern = "^T.*",replacement = "Theft",x=breach_report$Type.of.Breach)
 breach_report$Type.of.Breach<-sub(pattern = "^L.*",replacement = "Loss",x=breach_report$Type.of.Breach)
breach_report$Type.of.Breach<-sub(pattern = "^I.*",replacement = "Improper Use/Disposal",x=breach_report$Type.of.Breach)
breach_report$Type.of.Breach<-sub(pattern = "^I.*",replacement = "Improper Disposal",x=breach_report$Type.of.Breach)
 breach_report$Type.of.Breach<-sub(pattern = "^O.*",replacement = "Other",x=breach_report$Type.of.Breach)
breach_report$Type.of.Breach[breach_report$Type.of.Breach==""]<-"Unknown"
#following does what i want but how do i change colors, change order?
#stuck on the histogram piece can we talk through?
 ggplot(breach_report,aes(x=Type.of.Breach,fill=Individuals.Affected))+geom_bar()
 ggplot(breach_report,aes(x=Covered.Entity.Type,fill=Individuals.Affected))+geom_bar()
 # i want to do a series of histgrams...the example looks like i put the categorical variable on x but that gives me an error (and doesn't seem right for this)
ggplot(breach_report,aes(Individuals.Affected, fill=Type.of.Breach))+geom_histogram()


#ggplot(breach_report,aes( x=(Type.of.Breach), y=(Covered.Entity.Type)))+geom_point()
# this works and gets it totaled by day, how do i get it to be totaled by year?
 #aggregate(breach_reportIndividuals Affected`,by=list(breach_report$`Breach Submission Date`),sum)
#this totals all of the breach reports...how do i get it to total by type of breach?
breach_report %>% group_by('Type.of.Breach') %>% summarise(sum=sum(`Individuals.Affected`))
# totals by type of breach
#aggregate(breach_report$`Individuals.Affected`, by =list(Type=breach_report$`Type.of.Breach`),FUN=sum)

ggplot(breach_report,aes(Breach.Submission.Date,y=Individuals.Affected))+geom_line()


```
